Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2017-08-01 paper.tex
Add edits from 7.25.17 meeting

%\subsubsection*{In what ways is this method limited?}
Ray tracing is ideally suited to problems requiring detailed knowledge of
radiation distribution functions over small regions of spacetime:
for example along a matter or radiation trajectory,
or over a small volume outside a source.
But the method becomes computationally burdensome when the inquiry extends
to large volumes of spacetime.

@@

\subsubsection*{Outline}
In Sec.~\ref{sec:formulation} we derive the ray tracing equations from the
Boltzmann Equation and describe our numerical scheme.
In Sec.~\ref{sec:tests} we present tests of the code.
In Sec.~\ref{sec:applications} we present an application
in the dynamical environment following the merger of two neutron stars,
computing neutrino flavor oscillation along an outgoing ray,
including the effects of coherent forward scattering with ambient neutrinos.

@@

\subsection{Boltzmann Equation}
\label{ssec:boltzmann}
Neutrino radiation obeys the relativistic Boltzmann Equation,
which, suppressing the arguments of $f$ for simplicity, is written,
\todo{point out limit of QKEs}
\begin{equation}
  \label{eqn:boltzmann}
  \frac{d}{d\lambda}f = C[f],
\end{equation}
where $d/d\lambda$ denotes a derivative with respect to the affine
parameter defining the neutrino momentum (Eqn.~\ref{eqn:geodesic_x})
and $C[f]$ is the source term arising from interactions with the medium,
having dimension ${\rm energy}\,{\rm length}^{-1}$.
The source term varies over phase space $(x^\alpha,p_\beta)$,
and due to Fermi blocking depends locally on the distribution function $f$
and nonlocally on the distribution function of this neutrino and its
antiparticle at different momenta, $f'$ and $\bar{f}'$.
For simplicity we symbolize all of these dependencies with the shorthand $C[f]$.
The various neutrino interactions contributing to $C[f]$ are detailed in
App.~\ref{sec:source_terms}.

@@

These coefficients are computed by considering their dependence
on neutrino and antineutrino distribution functions at other momenta
(i.e.\ Fermi-blocking).
We consider two classes of interactions in this work:
the absorption/emission (AE) and elastic scattering (SE) processes
listed in Tab.~\ref{tab:neutrino_processes}.
Thus we separate these coefficients:
\begin{align}
  \label{eqn:inv_emissivity_components}
  \mathscr{E} &= \mathscr{E}_{\rm AE} + \mathscr{E}_{\rm SE},\\
  \label{eqn:inv_opacity_components}
  \mathscr{K} &= \mathscr{K}_{\rm AE} + \mathscr{K}_{\rm SE}.
\end{align}

The absorption/emission coefficients are computed from sums
over the relevant emissivities and opacities
in Tab.~\ref{tab:neutrino_processes}:
\todo{define $j$ and $\chi$}
\begin{align}
  \label{eqn:ae_emissivity_summed}
  \mathscr{E}_{\rm AE}(\varepsilon)
  &= \varepsilon \sum_{i\,{\rm reactions}} j_i(\varepsilon), \\
  \label{eqn:ae_opacity_summed}
  \mathscr{K}_{\rm AE}(\varepsilon)
  &= \frac{1}{1-f^{\rm eq}(\varepsilon)} \,
  \varepsilon \sum_{i\,{\rm reactions}} \chi_{{\rm a},i}(\varepsilon),
\end{align}
where $f^{\rm eq}$ is the distribution function of neutrinos in radiative
equilibrium with the matter, i.e. the Fermi-Dirac distribution function
\begin{equation}
  \label{eqn:feq}
  f^{\rm eq}(\varepsilon) \equiv
  \left(1+e^{\varepsilon/(k_{\rm B} T) -\eta_\nu}\right)^{-1},
\end{equation}
with the neutrino chemical potentials dependent on the local density,
temperature, and composition of the fluid via the neutron, proton,
and electron chemical potentials:
$\eta_{\nu_e}=-\eta_{\bar{\nu}_e}=\eta_p-\eta_n+\eta_{e^-}$ and
$\eta_{\nu_x}=0$.
The appearance of $f^{\rm eq}$ in Eqn.~\ref{eqn:ae_opacity_summed} is due to
the Fermionic nature of the neutrinos, causing $\mathscr{K}_{\rm AE}$ to be
different than the simple absorption opacity,
a phenomenon called stimulated absorption \cite{burr2006-neutrino_opacities}.
By detailed balance of the absorption/emission reactions, we may alternatively
write the emissivity in terms of the equilibrium distribution function:
\begin{equation}
  \label{eqn:ae_emissivity_feq}
  \mathscr{E}_{\rm AE}(\varepsilon)
  = \mathscr{K}_{\rm AE}(\varepsilon) f^{\rm eq}(\varepsilon).
\end{equation}
See App.~\ref{ssec:sources_ae} for details.
\todo{identify $\mathscr{K}_{\rm AE}$ and $\kappa^*$,
  in e.g.\ \cite{hari2010-gr_nunubar_collapsar}}

The elastic scattering coefficients are computed from
a background field, and a sum over opacities for
the relevant reactions in Tab.~\ref{tab:neutrino_processes}:
\begin{align}
  \label{eqn:se_emissivity}
  \mathscr{E}_{\rm SE}(\varepsilon)
  &= \mathscr{K}_{\rm SE}(\varepsilon) \, \Phi(\varepsilon) \, \\
  \label{eqn:se_opacity_summed}
  \mathscr{K}_{\rm SE}(\varepsilon)
  &= \varepsilon \sum_{i\,{\rm reactions}} \chi_{{\rm s},i}(\varepsilon).
\end{align}
See App.~\ref{ssec:sources_se} for details.

>>>>

%\subsubsection*{In what ways is this method limited?}
Ray tracing is ideally suited to problems requiring detailed knowledge of
radiation distribution functions over small regions of spacetime:
for example along a matter or radiation trajectory,
or over a small volume outside a source.

@@

In Sec.~\ref{sec:formulation} we derive the ray tracing equations from the
Boltzmann Equation and describe our numerical scheme.
In Sec.~\ref{sec:tests} we present tests of the code.
In Sec.~\ref{sec:applications} we present an application
in the dynamical environment following the merger of two neutron stars,
computing neutrino flavor oscillation along an outgoing ray,
including the effects of coherent forward scattering with ambient neutrinos.

@@

\subsection{Boltzmann Equation}
\label{ssec:boltzmann}
Neutrino radiation obeys the relativistic Boltzmann Equation,
which, suppressing the arguments of $f$ for simplicity, is written,
\todo{point out limit of QKEs}
\begin{equation}
  \label{eqn:boltzmann}
  \frac{d}{d\lambda}f = C[f],
\end{equation}
where $d/d\lambda$ denotes a derivative with respect to the affine
parameter defining the neutrino momentum (Eqn.~\ref{eqn:geodesic_x})
and $C[f]$ is the source term arising from interactions with the medium.
By Eqn.~\ref{eqn:geodesic_x},
the affine parameter has dimension ${\rm length}\,{\rm energy}^{-1}$,
so the source term has dimension ${\rm energy}\,{\rm length}^{-1}$.
The source term varies over phase space $(x^\alpha,p_\beta)$,
and depends locally on the distribution function $f$
and nonlocally on the distribution function of this neutrino and its
antiparticle at different momenta, $f'$ and $\bar{f}'$.
For simplicity we symbolize all of these dependencies with the shorthand $C[f]$.
The various neutrino interactions contributing to $C[f]$ are detailed in
App.~\ref{sec:source_terms}.

@@

These coefficients are computed by considering their dependence
on neutrino and antineutrino distribution functions at other momenta
(i.e.\ Fermi-blocking).
We consider the two dominant classes of interactions in this work:
the absorption/emission (AE) and elastic scattering (SE) processes
listed in Tab.~\ref{tab:neutrino_processes}.
Thus we separate these coefficients:
\begin{align}
  \label{eqn:inv_emissivity_components}
  \mathscr{E} &= \mathscr{E}_{\rm AE} + \mathscr{E}_{\rm SE},\\
  \label{eqn:inv_opacity_components}
  \mathscr{K} &= \mathscr{K}_{\rm AE} + \mathscr{K}_{\rm SE}.
\end{align}

The absorption/emission coefficients are computed from sums
over the relevant emissivities and opacities
for the reactions
\begin{align}
  \nu_e + n            &\leftrightarrow e^- + p,\nonumber\\
  \bar{\nu}_e + p      &\leftrightarrow e^+ + n,\nonumber\\
  \nu_e + {}^AZ        &\leftrightarrow e^- + {}^A(Z+1),\nonumber\\
  \bar{\nu}_e + {}^AZ  &\leftrightarrow e^+ + {}^A(Z-1),\nonumber
\end{align}
with ${}^ZA$ representing a nucleus of mass number $A$ and charge $Z$.
In terms of the emissivity $j(\varepsilon)$ describing number of neutrinos
of energy $\varepsilon$ emitted per length,
and the absorption opacity $\chi_a(\varepsilon)$
describing the number absorbed per length,
the coefficients are
\begin{align}
  \label{eqn:ae_emissivity_summed}
  \mathscr{E}_{\rm AE}(\varepsilon)
  &= \varepsilon \sum_{i\,{\rm reactions}} j_i(\varepsilon), \\
  \label{eqn:ae_opacity_summed}
  \mathscr{K}_{\rm AE}(\varepsilon)
  &= \frac{1}{1-f^{\rm eq}(\varepsilon)} \,
  \varepsilon \sum_{i\,{\rm reactions}} \chi_{{\rm a},i}(\varepsilon),
\end{align}
where $f^{\rm eq}$ is the distribution function of neutrinos in radiative
equilibrium with the matter, i.e. the Fermi-Dirac distribution function
\begin{equation}
  \label{eqn:feq}
  f^{\rm eq}(\varepsilon) \equiv
  \left(1+e^{\varepsilon/(k_{\rm B} T) -\eta_\nu}\right)^{-1},
\end{equation}
with the neutrino chemical potentials dependent on the local density,
temperature, and composition of the fluid via the neutron, proton,
and electron chemical potentials:
$\eta_{\nu_e}=-\eta_{\bar{\nu}_e}=\eta_p-\eta_n+\eta_{e^-}$ and
$\eta_{\nu_x}=0$.
The appearance of $f^{\rm eq}$ in Eqn.~\ref{eqn:ae_opacity_summed} is due to
the Fermionic nature of the neutrinos, causing $\mathscr{K}_{\rm AE}$ to be
different than the simple absorption opacity,
a phenomenon called stimulated absorption \cite{burr2006-neutrino_opacities}.
By detailed balance of the absorption/emission reactions, we may alternatively
write the emissivity in terms of the equilibrium distribution function:
\begin{equation}
  \label{eqn:ae_emissivity_feq}
  \mathscr{E}_{\rm AE}(\varepsilon)
  = \mathscr{K}_{\rm AE}(\varepsilon) f^{\rm eq}(\varepsilon).
\end{equation}
See App.~\ref{ssec:sources_ae} for details.
\todo{identify $\mathscr{K}_{\rm AE}$ and $\kappa^*$,
  in e.g.\ \cite{hari2010-gr_nunubar_collapsar}}

The elastic scattering coefficients are computed from
a background field, and a sum over opacities for
the reactions
\begin{align}
  \nu + N     &\leftrightarrow \nu + N,\nonumber\\
  \nu + {}^AZ &\leftrightarrow \nu + {}^AZ,\nonumber
\end{align}
with $N$ standing in for either $n$ or $p$.
In terms of a background field $\Phi(\varepsilon)$ describing
the number of neutrinos of energy $\varepsilon$ present at this event,
and the scattering opacity $\chi(\varepsilon)$ describing the
number scattered to other directions per length,
the coefficients are
\begin{align}
  \label{eqn:se_emissivity}
  \mathscr{E}_{\rm SE}(\varepsilon)
  &= \mathscr{K}_{\rm SE}(\varepsilon) \, \Phi(\varepsilon) \, \\
  \label{eqn:se_opacity_summed}
  \mathscr{K}_{\rm SE}(\varepsilon)
  &= \varepsilon \sum_{i\,{\rm reactions}} \chi_{{\rm s},i}(\varepsilon).
\end{align}
In the isotropic limit of trapped radiation,
$\Phi$ is equivalent to $f^{\rm eq}$;
in the free-streaming limit at a distance $r$ from a source,
$\Phi$ attenuates as $r^{-2}$.
See App.~\ref{ssec:sources_se} for details.
